---
title: "HW3_solutions"
author: "Dustin Pluta"
date: "April 25, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
```

## Problem 1

### b (i)

```{r}
#####
#####  Set the random number generator so that we all get the same results
#####

set.seed(111)


#####
#####  Set the simulation parameters
#####

n_sims <- 5000				## The total number of simulations we intend to run
n_grp <- c(30, 30, 30)			## The number of observations per group
grp_means <- c(1, 1, 1)		## The true mean of the distribution each group is sampled from
grp_var <- c(1, 1, 1)			## The true variance of each observation from each of the three groups		

#####
#####  Initialize the matrix that will hold the results of our tests 
#####  (test statistics will be in column 1; corresponding p-values will be in column 2) 
#####

result <- matrix(NA, nrow = n_sims, ncol = 2)

#####
#####  Run the simulation n_sims times 
#####
for(i in 1:n_sims){
	
	## Sample the data for each group, depending upon the specified mean and variance
	y <- rnorm(n = sum(n_grp), mean = rep(grp_means, times = n_grp), sd = rep(sqrt(grp_var), times = n_grp))
	group <- rep(1:3, times = n_grp)
	sim_data <- as.data.frame(cbind(group, y))
		
	## Fit the one-way ANOVA model, then extract the F-statistic and the p-value
	fit <- aov(y ~ factor(group), data = sim_data)
	F_stat <- summary(fit)[[1]][1, 4]
	p_val <- summary(fit)[[1]][1, 5]
	
	##  Store the results of simulation loop i in row i of the results matrix
	result[i, ] <- c(F_stat, p_val)
}
```

```{r, echo=TRUE}
#####
#####  Calculate the observed type I error rate
#####
  
n_reject <- sum(result[, 2] < .05 )	
sim_type1_error <- n_reject / n_sims
sim_type1_error	
```

```{r}
#####
#####  Plot the observed distribution of F-statistics vs. the theoretical distribution
#####

hist(result[, 1], xlab = "Simulated F-statistics", prob = T, ylim = c(0,1), breaks = 30)
lines(seq(0, 10, by = .1), df(seq( 0, 10, by = .1), df1 = 3 - 1, df2 = sum(n_grp) - 3))
```

### b (ii)

```{r}
#####
#####  Set the random number generator so that we all get the same results
#####

set.seed(111)


#####
#####  Set the simulation parameters
#####

n_sims <- 5000				## The total number of simulations we intend to run
n_grp <- c(30, 30, 30)			## The number of observations per group
grp_means <- c(1, 1, 1)		## The true mean of the distribution each group is sampled from
grp_var <- c(1, 1, 1)			## The true variance of each observation from each of the three groups		

#####
#####  Initialize the matrix that will hold the results of our tests 
#####  (test statistics will be in column 1; corresponding p-values will be in column 2) 
#####

result <- matrix(NA, nrow = n_sims, ncol = 2)

#####
#####  Run the simulation n_sims times 
#####
for(i in 1:n_sims){
	
	## Sample the data for each group, depending upon the specified mean and variance
	y <- rnorm(n = sum(n_grp), mean = rep(grp_means, times = n_grp), sd = rep(sqrt(grp_var), times = n_grp))
	group <- rep(1:3, times = n_grp)
	sim_data <- as.data.frame(cbind(group, y))
		
	## Fit the one-way ANOVA model, then extract the F-statistic and the p-value
	fit <- aov(y ~ factor(group), data = sim_data)
	F_stat <- summary(fit)[[1]][1, 4]
	p_val <- summary(fit)[[1]][1, 5]
	
	##  Store the results of simulation loop i in row i of the results matrix
	result[i, ] <- c(F_stat, p_val)
}
```

```{r, echo=TRUE}
#####
#####  Calculate the observed type I error rate
#####
  
n_reject <- sum(result[, 2] < .05 )	
sim_type1_error <- n_reject / n_sims
sim_type1_error	
```

```{r}
#####
#####  Plot the observed distribution of F-statistics vs. the theoretical distribution
#####

hist(result[, 1], xlab = "Simulated F-statistics", prob = T, ylim = c(0,1), breaks = 30)
lines(seq(0, 10, by = .1), df(seq( 0, 10, by = .1), df1 = 3 - 1, df2 = sum(n_grp) - 3))
```

### b (iii)

```{r}
#####
#####  Set the random number generator so that we all get the same results
#####

set.seed(111)


#####
#####  Set the simulation parameters
#####

n_sims <- 5000				## The total number of simulations we intend to run
n_grp <- c(30, 30, 30)			## The number of observations per group
grp_means <- c(1, 1, 1)		## The true mean of the distribution each group is sampled from
grp_var <- c(1, 1, 1)			## The true variance of each observation from each of the three groups		

#####
#####  Initialize the matrix that will hold the results of our tests 
#####  (test statistics will be in column 1; corresponding p-values will be in column 2) 
#####

result <- matrix(NA, nrow = n_sims, ncol = 2)

#####
#####  Run the simulation n_sims times 
#####
for(i in 1:n_sims){
	
	## Sample the data for each group, depending upon the specified mean and variance
	y <- rnorm(n = sum(n_grp), mean = rep(grp_means, times = n_grp), sd = rep(sqrt(grp_var), times = n_grp))
	group <- rep(1:3, times = n_grp)
	sim_data <- as.data.frame(cbind(group, y))
		
	## Fit the one-way ANOVA model, then extract the F-statistic and the p-value
	fit <- aov(y ~ factor(group), data = sim_data)
	F_stat <- summary(fit)[[1]][1, 4]
	p_val <- summary(fit)[[1]][1, 5]
	
	##  Store the results of simulation loop i in row i of the results matrix
	result[i, ] <- c(F_stat, p_val)
}
```

```{r, echo=TRUE}
#####
#####  Calculate the observed type I error rate
#####
  
n_reject <- sum(result[, 2] < .05 )	
sim_type1_error <- n_reject / n_sims
sim_type1_error	
```

```{r}
#####
#####  Plot the observed distribution of F-statistics vs. the theoretical distribution
#####

hist(result[, 1], xlab = "Simulated F-statistics", prob = T, ylim = c(0,1), breaks = 30)
lines(seq(0, 10, by = .1), df(seq( 0, 10, by = .1), df1 = 3 - 1, df2 = sum(n_grp) - 3))
```

\newpage 
## Problem 2

### (a)

Source     | SS      | df     | MS     | F
---------- | ------- | ------ | ------ | ---------
Treatment  | 4.66    | 2      | 2.33   | 1.674  
Error      | 197.64  | 142    | 1.392  | 
Total      | 202.3   | 144    |        | 

### (b)

```{r}
pf(1.674, 2, 142, lower.tail = FALSE)
```

\newpage
## Problem 3

### (a) & (b)

```{r, echo=FALSE}
# Simulating the Two-way ANOVA Model
###
# install.packages("dplyr")
# install.packages("ggplot2")
library(dplyr)
library(ggplot2)
###

# This function will return a data frame containing response values generated
# from a two-way ANOVA model with the specified parameters.
simulate_data <- function(mu, alpha, beta, delta, K, sigma) {
  I <- length(alpha)
  J <- length(beta)
  dat <- as.data.frame(expand.grid(i = 1:I, j = 1:J, k = 1:K, Y = NA))
  for (obs in 1:nrow(dat)) {
    dat$Y[obs] <- with(dat, mu + alpha[i[obs]] + beta[j[obs]] + delta[i[obs], j[obs]] + rnorm(1, 0, sigma))
  }
  return(dat)
}

###
# Simulate Data with no interactions
I <- 2 # Levels in Trt A
J <- 3 # Levels in Trt B
K <- 1000 # Obs per cell

mu <- 10
alpha <- c(-0.5, 0.5)
beta <- c(-1, 0, 1)
delta <- matrix(c(0, 0, 0, 0, 0, 0), nrow = length(alpha), ncol = length(beta))
sigma <- 1

dat <- simulate_data(mu, alpha, beta, delta, K, sigma)

dat_grp_means <- dat %>% group_by(i, j) %>% summarize(grp_mean = mean(Y))
```

The true interaction plot consists of two parallel lines.

```{r}
ggplot(dat_grp_means) + 
  geom_line(aes(x = factor(j), y = grp_mean, group = factor(i), linetype = factor(i), color = factor(i)), size = 2)
```

### (c)

The true deltas are 

$$\delta = \begin{pmatrix} -1 & 2 & -1\\ 1 & -2 & 1\end{pmatrix}.$$

```{r, echo=TRUE}
####
K <- 10000
mu <- 1
alpha <- c(-0.5, 0.5)
beta <- c(-1, 0, 1)
delta <- matrix(c(-1, 1, 2, -2, -1, 1), nrow = length(alpha), ncol = length(beta))
sigma <- 1

simulate_data <- function(mu, alpha, beta, delta, K, sigma) {
  I <- length(alpha)
  J <- length(beta)
  dat <- as.data.frame(expand.grid(i = 1:I, j = 1:J, k = 1:K, Y = NA))
  for (obs in 1:nrow(dat)) {
    dat$Y[obs] <- with(dat, mu + alpha[i[obs]] + beta[j[obs]] + delta[i[obs], j[obs]] + rnorm(1, 0, sigma))
  }
  return(dat)
}

dat <- simulate_data(mu, alpha, beta, delta, K, sigma)
dat_grp_means <- dat %>% group_by(i, j) %>% summarize(grp_mean = mean(Y))

ggplot(dat_grp_means) + 
  geom_line(aes(x = factor(j), y = grp_mean, group = factor(i), linetype = factor(i), color = factor(i)), size = 2)
```

\newpage
## Problem 4

### (a)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
dat_raw <-
read_csv(url("https://raw.githubusercontent.com/dspluta/STAT120C/master/Data/Chapter%2012/iron.csv"))
```

```{r}
dat <- dat_raw %>% 
  gather(key = "trt", value = "Y") %>% 
  mutate(form = stringr::str_split(trt, " ", simplify = TRUE)[, 1],
         dose = stringr::str_split(trt, " ", simplify = TRUE)[, 2])

# Alternative way
ggplot(dat) + 
  geom_boxplot(aes(x = form, y = Y, group = form, fill = form), alpha = 0.5) + 
  facet_grid(cols = vars(dose))
```

\newpage
### (b)


```{r}
fit <- lm(Y ~ form, data = dat)
knitr::kable(anova(fit))
```

### (c)

```{r}
fit <- lm(Y ~ factor(dose), data = dat)
knitr::kable(anova(fit))
```

### (d)

```{r}
fit <- lm(Y ~ form*factor(dose), data = dat)
knitr::kable(anova(fit))
```

\newpage

## Problem 5

### (a) See class notes.

### (b)

Yes, it is possible to have interactions with no main effects, although this requires a specific form of symmetry in the interaction plot due to the constraints $\sum_i \delta_{ij} = \sum_j \delta_{ij} = 0$.  For instance,

```{r, echo=TRUE}
####
K <- 10000
mu <- 1
alpha <- c(0, 0)
beta <- c(0, 0, 0)
delta <- matrix(c(-1, 1, 2, -2, -1, 1), nrow = length(alpha), ncol = length(beta))
sigma <- 1

simulate_data <- function(mu, alpha, beta, delta, K, sigma) {
  I <- length(alpha)
  J <- length(beta)
  dat <- as.data.frame(expand.grid(i = 1:I, j = 1:J, k = 1:K, Y = NA))
  for (obs in 1:nrow(dat)) {
    dat$Y[obs] <- with(dat, mu + alpha[i[obs]] + beta[j[obs]] + delta[i[obs], j[obs]] + rnorm(1, 0, sigma))
  }
  return(dat)
}

dat <- simulate_data(mu, alpha, beta, delta, K, sigma)
dat_grp_means <- dat %>% group_by(i, j) %>% summarize(grp_mean = mean(Y))

ggplot(dat_grp_means) + 
  geom_line(aes(x = factor(j), y = grp_mean, group = factor(i), linetype = factor(i), color = factor(i)), size = 2)
```


